---
title: "PS-5"
author: "Kush Patel"
date: "2023-02-18"
output: html_document
---

```{r Load Packages}
library(tidyverse)
library(ggpubr)
```

### Load and check data

#### 2)
```{r Load data}
gapminder <- read_delim("gapminder.csv.bz2")

num_rows <- nrow(gapminder)
num_cols <- ncol(gapminder)
```

The gapminder dataset has `r num_rows` rows and `r num_cols` columns. 

#### 3)
```{r Print sample}
gapminder %>%
  sample_n(7)
```

The random sample of 7 lines of data looks OK. It has 7 rows and 
25 columns, which is exactly what we would expect. 

### Descriptive Statistics 

#### 1)

```{r number of countries}
name_len <- length(unique(gapminder$name))
iso2_len <- length(unique(gapminder$iso2))
iso3_len <- length(unique(gapminder$iso3))
```

There are `r name_len` countries in the dataset. 

#### 2)

##### a)
```{r iso2-name comparison}
iso2_name <- gapminder %>%
  select(name, iso2) %>%
  group_by(iso2) %>%
  summarize(Name_Num = length(unique(name)))

iso2_name

for(i in 1:length(iso2_name$iso2)) {
  if(iso2_name$Name_Num[i] > 1) {
    print(iso2_name$iso2[i])
  }
}
```

There is only 1 country for each iso-2 code. The only place where we find more
than 1 values for an iso-2 code is when the code is NA. 

##### b)
```{r iso3-name comparison}
iso3_name <- gapminder %>%
  select(name, iso3) %>%
  group_by(name) %>%
  summarize(Iso3_Num = length(unique(iso3)))

iso3_name

for(i in 1:length(iso3_name$name)) {
  if(iso3_name$Iso3_Num[i] > 1) {
    print(iso3_name$name[i])
  }
}
```

There are no names that have more than one iso3-codes. The only place where
we find more than 1 values for a name is when the name is NA. 
However, it seems to be that the additional 4 values for iso3-codes come from
two missing values, CHANISL and NLD CURACAO. 

#### 3)
```{r Max and min years}
min_max_year <- gapminder %>%
  select(time) %>%
  filter(!(is.na(time))) %>%
  summarize(min_year = min(time), max_year = max(time))

min_max_year
```

The minimum year in the data is `r min_max_year$min_year` and the maximum year
is `r min_max_year$max_year`. 

### CO2 emissions

#### 1)
```{r Missing CO2 and CO2_PC values}
missing_CO2_values <- gapminder %>%
  group_by(time) %>%
  summarize(missing_CO2 = sum(is.na(co2)))

missing_CO2_PC_values <- gapminder %>%
  group_by(time) %>%
  summarize(missing_CO2_PC = sum(is.na(co2_PC)))

max_missing_value_CO2 <- missing_CO2_values %>%
  summarize(max_value = max(missing_CO2))

max_missing_value_CO2_PC <- missing_CO2_PC_values %>%
  summarize(max_value = max(missing_CO2_PC))

missing_CO2_values

missing_CO2_PC_values

i <- missing_CO2_values$missing_CO2 == max_missing_value_CO2$max_value
missing_CO2_values[i, ]

j <- missing_CO2_PC_values$missing_CO2_PC == max_missing_value_CO2_PC$max_value
missing_CO2_PC_values[j, ]

```

The years with the highest missing CO2 values were 2017, 2018, and 2019. 

The years with the highest missing CO2_PC values were 2017, 2018, and 2019.

#### 2)
```{r The USA}
usa <- gapminder %>%
  filter(iso3 == "USA") %>%
  filter(!(is.na(co2)))

ggplot(usa) + 
  geom_line(aes(x = usa$time, y = usa$co2)) +
  labs(x = "Time (Years)", y = "CO2 Emissions") +
  ggtitle("CO2 emissions over time (USA)")
```

From the graph, we see that the CO2 emissions for the USA initially increase
from 1960 to 1973(approx.), where they reached their first local maxima. After 
that, they decreases for an year or two before reaching their second maxima in 
1977(approx.). They reduces a little until 1982(approx.) before steadily 
increasing until 2005(approx.). After that, they steadily decrease until 2019. 

```{r China}
china <- gapminder %>%
  filter(iso3 == "CHN") %>%
  filter(!(is.na(co2)))

ggplot(china) + 
  geom_line(aes(x = china$time, y = china$co2)) +
  labs(x = "Time (Years)", y = "CO2 Emissions") +
  ggtitle("CO2 emissions over time (China)")
```

From the graph, we see that the CO2 emissions for China increase exponentially
over time from 1960 to 2014(approx.), before reducing until 2019. 

```{r India}
india <- gapminder %>%
  filter(iso3 == "IND") %>%
  filter(!(is.na(co2)))

ggplot(india) + 
  geom_line(aes(x = india$time, y = india$co2)) +
  labs(x = "Time (Years)", y = "CO2 Emissions") +
  ggtitle("CO2 emissions over time (India)")
```

From the graph, we see that CO2 emissions for India increase exponentially over
time from 1960 to 2019. 

```{r Indonesia}
indonesia <- gapminder %>%
  filter(iso3 == "IDN") %>%
  filter(!(is.na(co2)))

ggplot(indonesia) + 
  geom_line(aes(x = indonesia$time, y = indonesia$co2)) +
  labs(x = "Time (Years)", y = "CO2 Emissions") +
  ggtitle("CO2 emissions over time (Indonesia)")
```

From the graph, we see that CO2 emissions for Indonesia increase roughly 
exponentially over time from 1960 to 2012(approx.) before taking a sharp turn 
downwards from 2012 to 2015(approx.). After that they increase linearly until 
2019. 

```{r Iran}
iran <- gapminder %>%
  filter(iso3 == "IRN") %>%
  filter(!(is.na(co2)))

ggplot(iran) + 
  geom_line(aes(x = iran$time, y = iran$co2)) +
  labs(x = "Time (Years)", y = "CO2 Emissions") +
  ggtitle("CO2 emissions over time (Iran)")
```

From the graph, we see that CO2 emissions for Iran increase approximately 
linearly over time, with the exception of a little dip in 1979(approx.) and 
a little increase back up in 1996(approx.). 

#### 3)
```{r The USA CO2 per capita}
usa_PC <- gapminder %>%
  filter(iso3 == "USA") %>%
  filter(!(is.na(co2_PC)))

ggplot(usa_PC) + 
  geom_line(aes(x = usa_PC$time, y = usa_PC$co2_PC)) +
  labs(x = "Time (Years)", y = "CO2 Emissions Per Capital") +
  ggtitle("CO2 emissions per capita over time (USA)")
```

From the graph, we see that the CO2 emissions per capita for the USA has a 
bimodal distribution. The first maxima occurs around the year 1975 and the 
second occurs around the year 2000. 

```{r China CO2 per capital}
china_PC <- gapminder %>%
  filter(iso3 == "CHN") %>%
  filter(!(is.na(co2_PC)))

ggplot(china_PC) + 
  geom_line(aes(x = china_PC$time, y = china_PC$co2_PC)) +
  labs(x = "Time (Years)", y = "CO2 Emissions Per Capital") +
  ggtitle("CO2 emissions per capita over time (China)")
```

From the graph, we see that the CO2 emissions per capita for China decrease
initially, but then show a steady exponential increase until the year 
2012(approx.) before decreasing slightly. 

```{r India CO2 per capital}
india_PC <- gapminder %>%
  filter(iso3 == "IND") %>%
  filter(!(is.na(co2_PC)))

ggplot(india_PC) + 
  geom_line(aes(x = india_PC$time, y = india_PC$co2_PC)) +
  labs(x = "Time (Years)", y = "CO2 Emissions Per Capital") +
  ggtitle("CO2 emissions per capita over time (India)")
```

From the graph, we see that the CO2 emissions per capita for India increase
exponentially over time from 1960 to 2019. 

```{r Indonesia CO2 per capital}
indonesia_PC <- gapminder %>%
  filter(iso3 == "IDN") %>%
  filter(!(is.na(co2_PC)))

ggplot(indonesia_PC) + 
  geom_line(aes(x = indonesia_PC$time, y = indonesia_PC$co2_PC)) +
  labs(x = "Time (Years)", y = "CO2 Emissions Per Capital") +
  ggtitle("CO2 emissions per capita over time (Indonesia)")
```

From the graph, we see that the CO2 emissions per capita for Indonesia increase
approximately exponentially, with the exception of some troughs and crests 
along the way. The most significant dip occurs in 2013(approx.) before it 
increases back up until 2019. 

```{r Iran CO2 per capital}
iran_PC <- gapminder %>%
  filter(iso3 == "IRN") %>%
  filter(!(is.na(co2_PC)))

ggplot(iran_PC) + 
  geom_line(aes(x = iran_PC$time, y = iran_PC$co2_PC)) +
  labs(x = "Time (Years)", y = "CO2 Emissions Per Capital") +
  ggtitle("CO2 emissions per capita over time (Iran)")
```

From the graph, we see that the CO2 emissions per capita for Iran increases
initially from 1960 to 1975(approx.) before it decreases until 1981(approx.). 
After that it increases approximately linearly until 2019. 

#### 4)
```{r Average CO2 emisions per capita for continents}
avg_CO2_PC_continents <- gapminder %>%
  select(region, co2_PC) %>%
  filter(!(is.na(region)), !(is.na(co2_PC))) %>%
  group_by(region) %>%
  summarize(Avg_CO2_PC = mean(co2_PC))

avg_CO2_PC_continents
```

From the table, we see that the average CO2 emission per capita is highest for 
Europe with a value of 7.95 metric tons per capita. After Europe, Americas and 
Asia are pretty close to each other with a value of 6.45 and 6.21 metric tons 
per capita respectively. After them, Oceania has the highest CO2 emissions per
capita with a value of 4.38 metric tons per capita. Africa has the lowest CO2
emissions per capita with a value of 0.93 metric tons per capita. 

#### 5)
```{r Average CO2 per capita plot}
time_given <- c(1960, 2016)

gapminder %>%
  select(time, co2_PC, region) %>%
  filter(!(is.na(time)), !(is.na(co2_PC)), !(is.na(region)),
         time %in% time_given) %>%
  group_by(region, time) %>%
  summarize(Avg_CO2_PC = mean(co2_PC)) %>%
  ggplot(aes(x = region, y = Avg_CO2_PC, fill = factor(time))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Continent", y = "Average CO2 per capita")
```

#### 6)
```{r smallest and largest CO2 emitters}
largest <- gapminder %>%
  filter(!(is.na(region)),
         !(is.na(name)), 
         !(is.na(co2_PC))) %>%
  group_by(name, region) %>%
  summarize(total_CO2 = sum(co2_PC)) %>%
  group_by(region) %>%
  filter(rank(desc(total_CO2)) < 4) %>%
  arrange(region, desc(total_CO2))

largest

smallest <- gapminder %>%
  filter(!(is.na(region)),
         !(is.na(name)), 
         !(is.na(co2_PC))) %>%
  group_by(name, region) %>%
  summarize(total_CO2 = sum(co2_PC)) %>%
  group_by(region) %>%
  filter(rank(total_CO2) < 4) %>%
  arrange(region, total_CO2)

smallest
```

The tibbles generated give the three largest and the three smallest emitters
of CO2 per capita for each continent. 

### GDP per Capita
#### 1)

```{r gdp per capita vs life expectancy plot 1960}
gapminder %>%
  filter(!(is.na(name)),
         !(is.na(region)),
         !(is.na(GDP_PC)), 
         !(is.na(lifeExpectancy)),
         !(is.na(totalPopulation)),
         time == 1960) %>%
  group_by(name) %>%
  ggplot(aes(x = lifeExpectancy, y = GDP_PC, colour = region, size = totalPopulation)) +
  geom_point() +
  labs(x = "Life Expectancy", y = "GDP per capita")
```

From the graph, we see that the life expectancy and GDP per capita of a country
are exponentially related. 

#### 2)
```{r gdp per capita vs life expectancy plot 2019}
gapminder %>%
  filter(!(is.na(name)),
         !(is.na(region)),
         !(is.na(GDP_PC)), 
         !(is.na(lifeExpectancy)),
         !(is.na(totalPopulation)),
         time == 2019) %>%
  group_by(name) %>%
  ggplot(aes(x = lifeExpectancy, y = GDP_PC, colour = region, size = totalPopulation)) +
  geom_point() +
  labs(x = "Life Expactancy", y = "GDP per capita")
  
```

From the graph, we see that the life expectancy and GDP per capita of a country
are exponentially related. 

#### 3)
When comparing the two previous graphs, we see the following major changes in 
the world:

a) The population of some countries in Asia grew drastically in comparison to 
the populations of countries in other continents. 

b) The life expectancy of almost all the countries grew over the given time 
period. 

c) The GDP per capita of almost all the countries seem to have grown over the 
given time period as is seen by the increase in the scale used for the graph. 

#### 4)

```{r Average life expectancy of continents in 1960 and 2019}
life_expectancy_1960 <- gapminder %>%
  filter(!(is.na(region)), 
         !(is.na(lifeExpectancy)),
         time == 1960) %>%
  group_by(region) %>%
  summarize(LE = mean(lifeExpectancy))

life_expectancy_1960

life_expectancy_2019 <- gapminder %>%
  filter(!(is.na(region)), 
         !(is.na(lifeExpectancy)),
         time == 2019) %>%
  group_by(region) %>%
  summarize(LE = mean(lifeExpectancy))

life_expectancy_2019
```

Yes, the results in the tibbles fit with what we see in the figures. The life
expectancy of all the continents increased over time. 

#### 5)
```{r growth of life expectancy}
life_expectancy_growth <- gapminder %>%
  select(region, lifeExpectancy, time) %>%
  filter(!(is.na(region)), 
         !(is.na(lifeExpectancy)), 
         !(is.na(time))) %>%
  group_by(region) %>%
  arrange(time) %>%
  mutate(growth = lifeExpectancy - lag(lifeExpectancy)) %>%
  filter(!(is.na(growth))) %>%
  summarize(Avg_growth = mean(growth))

life_expectancy_growth
```

#### 6)

```{r Histograms for GDP per capital for 1960 and 2019}
histogram_1960 <- gapminder %>%
  filter(time == 1960, 
         !(is.na(GDP_PC))) %>%
  ggplot(aes(x = GDP_PC)) +
  geom_histogram() +
  labs(x = "GDP per capita 1960")

histogram_2019 <- gapminder %>%
  filter(time == 2019, 
         !(is.na(GDP_PC))) %>%
  ggplot(aes(x = GDP_PC)) +
  geom_histogram() +
  labs(x = "GDP per capita 2019")

histograms <- ggarrange(histogram_1960, histogram_2019, ncol = 2, nrow = 1)

histograms
```

#### 7)

```{r US rank}
rankings_1960 <- gapminder %>%
  select(name, time, lifeExpectancy) %>%
  filter(!(is.na(name)), 
         !(is.na(lifeExpectancy)), 
         time == 1960) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  arrange(rank)

rankings_1960

rankings_2019 <- gapminder %>%
  select(name, time, lifeExpectancy) %>%
  filter(!(is.na(name)), 
         !(is.na(lifeExpectancy)), 
         time == 2019) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  arrange(rank)

rankings_2019
```

The United States was ranked 17th in terms of life expectancy in 1960 and 46th
in 2019. 

#### 8)
```{r relative rank}
num_countries_1960 <- gapminder %>%
  filter(time == 1960,
         !(is.na(iso3))) %>%
  summarize(num_countries = length(unique(iso3)))

num_countries_2019 <- gapminder %>%
  filter(time == 2019,
         !(is.na(iso3))) %>%
  summarize(num_countries = length(unique(iso3)))

relative_1960_US <- 17/num_countries_1960$num_countries
  
relative_2019_US <- 46/num_countries_2019$num_countries

relative_1960_US
relative_2019_US
```

The US was relatively ranked `r relative_1960_US` in term of life expectancy in
1960. 

The US was relatively ranked `r relative_2019_US` in term of life expectancy in
2019. 

### How many hours did you spend on this PS.

I spent 6 hours on this PS. 